---
title: "How do player performance statistics affect the outcome of a Big Ten Conference baseball game?"
author: "Kyle Gilbert"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: lualatex
    number_sections: true
---

```{r Imports, warning=FALSE, message=FALSE, echo=FALSE}

# Import packages
library(dplyr)
library(tidyr)
library(purrr)
library(kableExtra)
library(knitr)
library(tinytex)
library(stats)
library(stringr)
library(betareg)
library(pROC)
library(lubridate)
library(caret)
library(glue)
library(car)

# Load data
files <- list.files("~/Github/big-ten-baseball-model/data/", full.names = TRUE)
invisible(lapply(files, load, envir = .GlobalEnv))
```

```{r Globals, echo=FALSE}

# List teams in the big ten conference
bigten.teams <- c("ILL_ILL", "IU", "IOW_HAW", "MAR_TER", "MIC_SPA", "MIC_WOL", 
  "MIN_GOL", "NEB", "NOR_CAT", "OSU_BUC", "PEN_NIT", "PUR_BOI", "RUT_SCA")

# Specify factors for frequency analysis
factors <- list("PlayResult", "KorBB", "PitchCall", "RunsScored")

# Specify grouping variables
variables <- c("GameID", "HomeTeam", "AwayTeam", "Date")

# Specify predictor variable types
pred.vars <- c("FIP", "WHIP", "wOBA", "ISO", "K9", "BB9", "K%", "BB%")
```


```{r Data Cleaning, warning=FALSE, echo=FALSE}

# Loop through global environment
for (name in ls()) {
  
  # Get environment objects
  object <- get(name)
  
  # Get data frame objects
  if (is.data.frame(object) & grepl(paste(bigten.teams, collapse = "|"), name)) {
    
    # Remove "undefined" factor level
    levels(object$PlayResult)[levels(object$PlayResult) == "Undefined"] <- NA
    levels(object$KorBB)[levels(object$KorBB) == "Undefined"] <- NA
    
    # Filter in-conference game data
    temp <- subset(object, HomeTeam %in% bigten.teams & AwayTeam %in% bigten.teams)
    assign(name, temp)
  }
}

# Remove temporary variables
rm(name, files, temp, object)
```

```{r Transformation Function, echo=FALSE}

# Define frequency analysis function
freq <- function(data, type) {
  map(factors, ~ data %>% 
    count(GameID, {{type}}, .data[[.x]]) %>% 
    pivot_wider(
      names_from = .data[[.x]],
      values_from = n,
      values_fill = 0
    )
  )
}

# Define function to summarize team data
summarize.team <- function(data) {
  
  # Frequency analysis for factors
  pitcher.counts <- freq(data, Pitcher)
  batter.counts <- freq(data, Batter)
  
  # Derive innings-pitched pitcher statistic
  IP <- data %>% distinct(!!!syms(variables), Inning, Pitcher, PitcherTeam) %>% 
    count(!!!syms(variables), Date, Pitcher, PitcherTeam, name = "IP")
  
  # Derive plate appearances batting statistic
  PA <- data %>% distinct(!!!syms(variables), Inning, PAofInning, Batter, BatterTeam) %>%
    count(!!!syms(variables), Date, Batter, BatterTeam, name = "PA")
  
  # Specify variables to join by
  keys <- c("GameID", "Pitcher")
  
  # Join pitcher counts and innings pitched
  pitcher.data <- left_join(IP, reduce(pitcher.counts, left_join, by = keys), by = keys) %>% 
    mutate(
      
      # Fielding independent pitching (FIP)
      FIP = (13*HomeRun + 3*(Walk + HitByPitch) - 2*Strikeout) / IP + 4.22,
      
      # Strikeouts & walks per nine innings (K/9, BB/9)
      K9 = (Strikeout / IP) * 9, BB9 = (Walk / IP) * 9,
      
      # Walks & hits per inning pitched
      WHIP = (Walk + Single + Double + 
        tryCatch({Triple}, error = function(cond) {0}) + HomeRun) / IP
      
    ) %>% 
    
    # Summarize across both teams
    reframe(
      Date = Date,
      HomeTeam = HomeTeam,
      AwayTeam = AwayTeam,
      across(
        c(FIP, K9, BB9, WHIP),
        ~mean(.x[PitcherTeam == HomeTeam]),
        .names = "Home_{.col}"),
      across(
        c(FIP, K9, BB9, WHIP),
        ~mean(.x[PitcherTeam == AwayTeam]),
        .names = "Away_{.col}"),
      .by = GameID
    )
  
  # Update keys for batter data
  keys <- c("GameID", "Batter")
  
  # Join batter counts and plate appearances
  batter.data <- left_join(PA, reduce(batter.counts, left_join, by = keys), by = keys) %>% 
    mutate(
      
      # Weighted on-base average (wOBA)
      wOBA = (0.812*Walk + 0.838*HitByPitch + 0.943*Single + 1.245*Double + 
        1.537*tryCatch({Triple}, error = function(cond) {0}) + 1.764*HomeRun) 
        / (PA + Walk + Sacrifice + HitByPitch),
      
      # Strikeout & walk percentage (K%, BB%)
      `K%` = Strikeout / PA, `BB%` = Walk / PA,
      
      # Isolated power (ISO)
      ISO = (Double + 2 * tryCatch({Triple}, error = function(cond) {0}) + 
        3 * HomeRun) / PA
      
    ) %>% 

    # Summarize across both teams
    reframe(
      Date = Date,
      HomeTeam = HomeTeam,
      AwayTeam = AwayTeam,
      across(
        c(wOBA, `K%`, `BB%`, ISO),
        ~mean(.x[BatterTeam == HomeTeam]),
        .names = "Home_{.col}"),
      across(
        c(wOBA, `K%`, `BB%`, ISO),
        ~mean(.x[BatterTeam == AwayTeam]),
        .names = "Away_{.col}"),
      .by = GameID
    )
  
  # Join batter and pitcher summary data
  pitcher.data %>% left_join(batter.data, by = "GameID") %>% distinct() %>% select(-GameID)
}
```

```{r Data Transformation, warning=FALSE, echo=FALSE}

# Summarize each team's in-conference game data
bigten.team.data <- lapply(bigten.teams, function(team) {summarize.team(get(team))})

# Merge in-conference game data for all teams
bigten.games <- Reduce(rbind, bigten.team.data) %>% 
  mutate(
    Date = coalesce(Date.x, Date.y),
    HomeTeam = coalesce(HomeTeam.x, HomeTeam.y),
    AwayTeam = coalesce(AwayTeam.x, AwayTeam.y)
  ) %>% 
  select(-Date.x, -Date.y, -HomeTeam.x, -HomeTeam.y, -AwayTeam.x, -AwayTeam.y) %>% 
  distinct()

# Remove incorrectly-dated game
bigten.games <- bigten.games[-61, ]

# Create 'result' column to store game result
# Result stores 1 if the home team wins, otherwise 0
bigten.games$Result <- c(0,0,1,0,1,1,1,1,1,0,1,1,1,0,1,1,0,1,1,1,0,1,0,0,0,1,0,
  1,1,0,1,0,1,1,1,1,0,0,0,1,1,0,1,1,0,0,0,1,1,1,0,0,1,0,1,1,1,1,1,1,0,1,1,1,0,1,
  0,1,1,1,1,0,1,0,1,0,1,0,0,0,0,0,0,1,1,1,1,1,1,1,0,1,1,0,1,1,0,0,1,1,0,0,1,0,0,
  1,0,1,1,0,1,0,1,1,1,0,0,0,0,1,1,0,0,0,1,1,0,1,0,1,1,1,1,0,0,1,1,0,1,1,0,0,0,1,
  1,0,1,0,0,0)
```

# Introduction

# Exploratory Data Analysis

## Variable Description

The response variable represents the outcome of a Big Ten Conference baseball game. As an indicator variable, the response value `1` indicates the home team won the game, whereas a response value of `0` means the away team emerged victorious. The predictors comprise 8 player performance statistics for each team, which amounts to 16 continuous explanatory variables. Refer to Table 1 for additional details.

Fielding Independent Pitching (FIP) quantifies a pitcher's performance solely based on the aspects of the game a pitcher controls, which are determined to be home runs (HR), walks (BB), hit-by-pitches (HBP), and strikeouts (K). The following formula for calculating FIP includes a constant that scales this statistic to that of Earned Run Average (ERA). 
$$
FIP = \frac{13\times HR + 3(BB + HBP) - 2 \times K}{IP} + constant
$$
Walks plus Hits per Inning Pitched (WHIP) measures a pitcher's ability to prevent hitters from reaching base, as base runners pose a scoring threat. The WHIP formula below accounts for pitchers allowing a base runner via a walk, single (1B), double (2B), triple (3B), or home run.
$$
WHIP = \frac{BB + 1B + 2B + 3B + HR}{IP}
$$
Strikeouts per Nine Innings (K/9) is the average number of strikeouts a pitcher has for every nine innings pitched. This rate statistic offers insight into the pitcher's ability to strike batters out; however, relief pitchers often have a higher K/9 than starting pitchers because relievers throw for a shorter duration, which enables them to exert more force on each pitch.
$$
K/9 = \frac{K}{IP} \times 9
$$
Walks per Nine Innings (BB/9) is the average number of walks a pitcher gives up for every nine innings pitched. This rate statistic captures how often a pitcher allows a base runner through a walk.
$$
BB/9 = \frac{BB}{IP} \times 9
$$
Weighted On-Base Average (wOBA) evaluates a batter's performance by weighting the batting events that result in a batter reaching base with the 2024 season's adjusted run expectancy for each event.
$$
wOBA = \frac{0.812 \times BB + 0.838 \times HBP + 0.943 \times 1B + 1.245 \times 2B + 1.537 \times 3B + 1.764 \times HR}{AB + SF + BB + HBP}
$$
Isolated Power (ISO) considers extra-base hits to express a batter's power, as hits longer than a double signify powerful hits. The ISO formula gives more weight to longer extra-base hits and averages these hits over the number of at-bats.
$$
ISO = \frac{2B + 2 \times 3B + 3 \times HR}{AB}
$$
Strikeout Percentage (K%) is the frequency at which a batter strikes out, whereas Walk Percentage (BB%) is the frequency at which a batter earns a walk. These rate statistics assess a hitter's plate discipline.

$$
K\% = \frac{K}{AB} \qquad \qquad BB\% = \frac{BB}{AB}
$$

## Data Visualization

To further develop a comprehensive understanding of the data, we generate several visualizations to identify patterns in the data and examine potential relationships between the response and predictor variables prior to modeling.

```{r Table 1, echo=FALSE}

vars <- data.frame(
  
  # Variable names
  Variable = c(
    "Result", "Home.FIP", "Away.FIP", "Home.WHIP", "Away.WHIP", "Home.K9",
    "Away.K9", "Home.BB9", "Away.BB9", "Home.wOBA", "Away.wOBA", "Home.ISO",
    "Away.ISO", "Home.K%", "Away.K%", "Home.BB%", "Away.BB%"
  ),
  
  # Variable type in the model
  Type = c("Response", rep("Predictor", 16)),
  
  # An explanation of the variable
  Explanation = c(
    "An indicator variable representing whether the home team (1) or the away team (0) won the game",
    "Home team Fielding Independent Pitching (FIP) pitching statistic from the game",
    "Away team Fielding Independent Pitching (FIP) pitching statistic from the game",
    "Home team Walks + Hits per Inning Pitched (WHIP) pitching statistic from the game",
    "Away team Walks + Hits per Inning Pitched (WHIP) pitching statistic from the game",
    "Home team pitcher Strikeouts per Nine Innings (K9) rate from the game",
    "Away team pitcher Strikeouts per Nine Innings (K9) rate from the game",
    "Home team pitcher Walks per Nine Innings (BB9) rate from the game",
    "Away team pitcher Walks per Nine Innings (BB9) rate from the game",
    "Home team Weighted On-Base Average (wOBA) batting statistic from the game",
    "Away team Weighted On-Base Average (wOBA) batting statistic from the game",
    "Home team Isolated Power (ISO) batting statistic from the game",
    "Away team Isolated Power (ISO) batting statistic from the game",
    "Home team batter Strikeout Percentage (K%) from the game",
    "Away team batter Strikeout Percentage (K%) from the game",
    "Home team batter Walk Percentage (BB%) from the game",
    "Away team batter Walk Percentage (BB%) from the game"
  ),
  
  # Range of values for each variable
  Value = c(
    glue("0: {round(sum(bigten.games$Result) / nrow(bigten.games),2)*100}% 
    1: {round(1 - (sum(bigten.games$Result) / nrow(bigten.games)),2)*100}%"),
    glue("{round(min(bigten.games$Home_FIP), 2)} ~ 
      {round(max(bigten.games$Home_FIP), 2)}"),
    glue("{round(min(bigten.games$Away_FIP), 2)} ~ 
      {round(max(bigten.games$Away_FIP), 2)}"),
    glue("{round(min(bigten.games$Home_WHIP), 2)} ~ 
      {round(max(bigten.games$Home_WHIP), 2)}"),
    glue("{round(min(bigten.games$Away_WHIP), 2)} ~ 
      {round(max(bigten.games$Away_WHIP), 2)}"),
    glue("{round(min(bigten.games$Home_K9), 2)} ~ 
      {round(max(bigten.games$Home_K9), 2)}"),
    glue("{round(min(bigten.games$Away_K9), 2)} ~ 
      {round(max(bigten.games$Away_K9), 2)}"),
    glue("{round(min(bigten.games$Home_BB9), 2)} ~ 
      {round(max(bigten.games$Home_BB9), 2)}"),
    glue("{round(min(bigten.games$Away_BB9), 2)} ~ 
      {round(max(bigten.games$Away_BB9), 2)}"),
    glue("{round(min(bigten.games$Home_wOBA), 2)} ~ 
      {round(max(bigten.games$Home_wOBA), 2)}"),
    glue("{round(min(bigten.games$Away_wOBA), 2)} ~ 
      {round(max(bigten.games$Away_wOBA), 2)}"),
    glue("{round(min(bigten.games$Home_ISO), 2)} ~ 
      {round(max(bigten.games$Home_ISO), 2)}"),
    glue("{round(min(bigten.games$Away_ISO), 2)} ~ 
      {round(max(bigten.games$Away_ISO), 2)}"),
    glue("{round(min(bigten.games$`Home_K%`), 2)} ~ 
      {round(max(bigten.games$`Home_K%`), 2)}"),
    glue("{round(min(bigten.games$`Away_K%`), 2)} ~ 
      {round(max(bigten.games$`Away_K%`), 2)}"),
    glue("{round(min(bigten.games$`Home_BB%`), 2)} ~ 
      {round(max(bigten.games$`Home_BB%`), 2)}"),
    glue("{round(min(bigten.games$`Away_BB%`), 2)} ~ 
      {round(max(bigten.games$`Away_BB%`), 2)}")
  )
)

vars %>% 
  knitr::kable(
    caption = "Variables used in the analysis",
    align = "l",
  ) %>% 
  kableExtra::kable_styling(
    position = "center",
    bootstrap_options = c("striped")
  ) %>% 
  kable_classic(
    html_font = "Cambria",
    full_width = FALSE
  ) %>% 
  column_spec(3, width = "8cm")
```

### Univariate Analysis

Before categorizing the player performance data into either the home team or away team data, Figure 1 displays the histogram of each predictor type. Although most graphs reveal an approximately symmetric distribution, the histogram showing Isolated Power (ISO) frequencies appears right-skewed. Therefore, a transformation of the ISO variable is needed to achieve normalization.

```{r Univariate Analysis, echo=FALSE, fig.height=9, fig.width=9}

# Define plot parameters
par(mfrow = c(3,3))

# Visualize predictor variables
for (var in pred.vars) {
  
  # Combine home and away statistic values
  data <- unlist(bigten.games[,grep(var, names(bigten.games), value = TRUE)])
  
  # Create a histogram from the data
  hist(data, main = paste("Histogram of", var), xlab = paste(var, "values"))
}
```

\begin{figure}[ht]
\caption{Histogram of each predictor type. Refer to Table 1 for variable name meanings.}
\end{figure}

The `Value` column in Table 1 conveys the range of values for the Isolated Power (ISO) variables. Notably, both the home and away ISO contain zero values. As a result, common transformations for remedying skewed data, like Box-Cox, are unfeasible because they require entirely positive data. We chose to apply a Yeo-Johnson transformation to the ISO variables to make the distribution of their values more normal. The Yeo-Johnson transformation is the outcome of augmenting the Box-Cox transformation to handle non-positive values, which means it can take the ISO values.

```{r Yeo-Johnson Transformation, echo=FALSE}

# Define plot parameters
par(mfrow = c(1,2))

# Combine home and away ISO values
ISO.before <- unlist(bigten.games[,grep("ISO", names(bigten.games), value = TRUE)]) %>% unname()

# Estimate optimal lambda value
powt <- powerTransform(ISO.before ~ 1, family = "yjPower")

# Apply Yeo-Johnson transformation
ISO.after <- yjPower(ISO.before, lambda = powt$lambda)
bigten.games$Home_ISO <- yjPower(bigten.games$Home_ISO, lambda = powt$lambda)
bigten.games$Away_ISO <- yjPower(bigten.games$Away_ISO, lambda = powt$lambda)

# ISO histogram before transformation
hist(ISO.before, main = "ISO before Yeo-Johnson", xlab = "ISO values")

# ISO histogram after transformation
hist(ISO.after, main = "ISO after Yeo-Johnson", xlab = "Transformed ISO values")
```


```{r Modeling, message=FALSE, warning=FALSE, echo=FALSE}

# Set seed for reproducibility
set.seed(123)

# Perform 70/30 train-test split
indices <- sample(nrow(bigten.games), size = 0.7 * nrow(bigten.games))
train <- bigten.games[indices,]
test <- bigten.games[-indices,]

# Define model predictors
model.formula <- formula(Result ~ Home_FIP + Away_FIP + Home_WHIP + Away_WHIP + 
  Home_K9 + Away_K9 + Home_BB9 + Away_BB9 + Home_wOBA + Away_wOBA + Home_ISO + 
  Away_ISO + `Home_K%` + `Away_K%` + `Home_BB%` + `Away_BB%`)

# Multiple linear regression model
linear.model <- lm(model.formula, data = train)

# Factorize 'result' response variable
train$Result <- factor(train$Result)

# Logistic regression model
logistic.model <- glm(model.formula, family = binomial, data = train)

# Predict test results
linear.pred <- predict(linear.model, test[,1:16], type = "response") %>% round(0)
logistic.pred <- predict(logistic.model, test[,1:16], type = "response") %>% round(0)

# Create confusion matrices
linear.conf <- confusionMatrix(as.factor(linear.pred), as.factor(test$Result))
logistic.conf <- confusionMatrix(as.factor(logistic.pred), as.factor(test$Result))

linear.conf$overall["Accuracy"]
logistic.conf$overall["Accuracy"]

rm(indices, size, train, vars, model, i)
rm(test, logistic.model, logistic.pred)
```

```{r Table 2, echo=FALSE}

model.summary <- data.frame(
  Model = c("Linear Regression", "Logistic Regression")
  
)
```

